A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2020-01-31 paper/main.tex
fixes typos and table with r squares

For the purposes of this manuscript a data set containing results of IPD
tournaments has been generated and is available at~\cite{data}. This was done using the
open source package Axelrod-Python library~\cite{axelrodproject} (APL), and more specifically,
version 3.0.0. APL allows for different types of IPD computer
tournaments to be simulated whilst containing a list of over 180 strategies.
Most of these are strategies described in the literature with a few exceptions
being strategies that have been contributed specifically to the package. This
paper make use of \numberofstrategies strategies implemented in version 3.0.0. A
list of the strategies is given in the Appendix~\ref{app:list_of_players}.
Although APL features several tournament types, this work considers
standard, noisy, probabilistic ending and noisy probabilistic ending
tournaments.

@@

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014}, which are
strategies based on finite state automata introduced by Daniel and Wendy
Ashlock. These strategies have been evolved using reinforcement learning, however,
there were trained to maximise their payoffs in tournaments with fixed turns
(150 specifically) and not in probabilistic ending ones. In probabilistic ending
tournaments it appears that the top ranks are mostly occupied by defecting
strategies. These include Better and Better, Gradual Killer, Hard Prober (all
from ~\cite{prison}), Bully (Reverse Tit For Tat)~\cite{Nachbar1992} and
Defector. Thus, it's surprisingly that EasyGo and Fool Me Forever which are
strategies that will defect until their opponent defect, then they will cooperate
until the end, ranked $14^{\text{th}}$ and $15^{\text{th}}$. Upon inspection,
it was found that they are actually the same strategy. This was not known to
the authors at the time of data collection. Figure~\ref{fig:probend_results}
verifies that their performance is the same. Both strategies have repeatedly
ranked highly and there are cases for which they were the winners of the
tournament.

@@

The top ranks include strategies that have been previously mentioned. The
set of Retaliate strategies occupy the top spots followed by BackStabber
and DoubleCrosser. The distributions of the Retaliate strategies have no
statistical difference. Thus, in an IPD tournament where the type is not
specified, playing as any of the Retaliate strategies will have the result.
DoubleCrosser performed well in standard tournaments and the
strategy is just an extension of BackStabber. It should be noted that these
strategies can be characterised as "cheaters". The source code of the strategies
allows them to known the number of turns in a match (if they are specified).
PSO Gambler and Evolved HMM 5 are
trained strategies introduced in~\cite{Harper2017} and Nice Meta Winner and NMWE
Memory One are strategies based on teams. Grudger is a strategy from Axelrod's
original tournament and Forgetful Fool Me Once is based on the same approach as
Grudger. Overall the top 15 strategies are fundamentally different. Some are cheaters,
some are complex, others are simple deterministic strategies and strategies based
on teams. The results of \numberofalltournaments tournaments used in this work imply the following:
they is not a single type of strategy which can performance well in any IPD interaction.

@@

Compared to standard tournaments, in both noisy and in probabilistic ending
tournaments the higher the rates of cooperation the lower a strategy's success
and median score. A strategy would want to cooperate less than both
the mean and median cooperator in such settings. In probabilistic ending
tournaments the correlations coefficients have a larger values, indicating a
stronger effect. Thus a strategy will be punished more by it's cooperative
behaviour in probabilistic ending environments, this was seen in Section~\ref{section:evaluation_of_performance}
as well. The distributions of the $C_r$ of the winners in
both tournaments is given by Figure~\ref{fig:c_r_distributions}. It confirms
that the winners in noisy tournaments cooperated less than 35\% of the times
and in probabilistic ending tournaments less than 10\%.
In noisy probabilistic ending tournaments and in over all the tournaments' results,
the only features that had a moderate affect are $C_r/C_{\text{mean}},
C_r/C_{\text{max}}$ and $C_r$. In such environments cooperative behaviour
appears to be punished by not as much as in noisy and probabilistic ending
tournaments.

@@

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.9\textwidth}{!}{
        \begin{tabular}{lccccc}
    \toprule
    Tournament type & Clustering Approach & Number of clusters & $R^2$ training data &  $R^2$ test data  & $R^2$ OOB score\\
    \midrule
    standard  & Approach 1 & 2  & 0.998831  & 0.987041 & 0.983708 \\
              & Approach 2 & 2  & 0.998643  & 0.978626 & 0.969202 \\
              & Approach 3 & 2  & 0.998417  & 0.985217 & 0.976538 \\
              & Approach 4 & 2  & 0.998794  & 0.990677 & 0.982959 \\
    \midrule
    noisy     & Approach 1  & 2 & 0.996677  & 0.950572 & 0.935383\\
              & Approach 2  & 2 & 0.996677  & 0.950572 & 0.935383\\
              & Approach 3  & 2 & 0.996677  & 0.950572 & 0.935383\\
              & Approach 4  & 3 & 0.996677  & 0.950572 & 0.935383\\
    \midrule
    probabilistic ending & Approach 1  & 2 & 0.999592  & 0.995128 & 0.992819 \\
                         & Approach 2  & 2 & 0.999592  & 0.995128 & 0.992819 \\
                         & Approach 3  & 2 & 0.999592  & 0.995128 & 0.992819 \\
                         & Approach 4  & 2 & 0.999592  & 0.995128 & 0.992819 \\
    \midrule
    noisy probabilistic ending & Approach 1  & 2 & 0.990490  & 0.813905 & 0.791418\\
                               & Approach 2  & 2 & 0.990490  & 0.813905 & 0.791418\\
                               & Approach 3  & 2 & 0.990490  & 0.813905 & 0.791418\\
                               & Approach 4  & 4 & 0.990490  & 0.813905 & 0.791418\\
    \midrule
    over \numberofalltournaments tournaments & Approach 1  & 2 & 0.993396 & 0.913409 & 0.898059 \\
                                             & Approach 2  & 2 & 0.993396 & 0.913409 & 0.898059 \\
                                             & Approach 3  & 2 & 0.993396 & 0.913409 & 0.898059 \\
                                             & Approach 4  & 3 & 0.993396 & 0.913409 & 0.898059 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Accuracy metrics for random forest models.}
    \label{table:accuracy_random_forest}
\end{table}

@@

The \numberofstrategies strategies used in this manuscript have been mainly for
the literature, and they have been accessible due to an open source software
called the Axelrod-Python library. The software was used to generate a total of
\numberofalltournaments computer tournaments results with different number of
strategies and different participants each time. The data collection was
described in Section~\ref{section:data_collection}. In
Section~\ref{section:top_performances}, the tournaments results were used to
present the top performances. The data set contained results from four different
settings, and these were also studied individually. In standard tournaments
complex strategies trained using reinforcement learning ranked in the top spots.
Some of these strategies ranked again in the top spots in probabilistic
ending tournaments when a \(p_e\) of less 0.1 was considered. In probabilistic
ending tournaments \(p_e\) was designed to vary between 0 and 1. It was demonstrated
that for values larger than 0.1, as stated in the Folk Theorem, defecting strategies
were winning the tournaments because there was a high likelihood of the game
ending in the next turn. In tournaments with noise the median ranks of the top
15 strategies had the highest values and the \(r\) distributions were bimodal.
The top rank strategies were performing both well and bad, and this indicates
that in noisy tournaments there are not strategies that can guarantee winning.
Overall, the top ranked strategies differed from one tournament type to
another and the mechanism behind the winning strategies were all different.
Even strategies designed to do good in one setting did better in others.
On the whole ... (the ipd interactions are unique and there is no winning
strategy) %TODO help from Vince

Section~\ref{section:evaluation_of_performance}, covered an analysis of
performance based on several features associated with a strategy and with the
environments it was competing. The results of this analysis showed that a
strategy's characteristics such as whether or not it's stochastic, and the information it
used regarding the game had no effect on the strategy's success. The most
important features have been those that compared the strategy's behaviour to it's
environment. The cooperating ratio of the strategy compared to the mean and
median cooperator was highlighted as the most important feature in the analysis.
More specifically, if a strategy were to enter a tournament with a theory of
mind of its environment it would choose to be the median cooperator in standard
tournaments, to cooperate 10\% of the time the median cooperator did in probabilistic ending tournaments and 
60\% in noisy and noisy probabilistic
tournaments. Lastly, if a strategy was aware of the opponents but not of the
setting of the tournament, a strategy would be more likely to be successful if
it were to identify the median cooperator and cooperated 67\% of the times that
they did.

@@

\begin{table}[h]
    \begin{center}
    \resizebox{.7\textwidth}{!}{
    \begin{tabular}{llc}
    \toprule
feature & feature explanation \\
 \midrule
stochastic                  & If a strategy is stochastic \\
makes use of game           & If a strategy makes used of the game information  \\
makes use of length         & If a strategy makes used of the number of turns \\
memory usage                & The memory size of a strategy divided by the number of turns \\
SSE                         & A measure of how far a strategy is from extortionate behaviour \\
$C_{\text{max}}$          & The biggest cooperating rate in the tournament \\
$C_{\text{min}}$          & The smallest cooperating rate in the tournament \\
$C_{\text{median}}$       & The median cooperating rate in the tournament \\
$C_{\text{mean}}$         & The mean cooperating rate in the tournament \\
$C_r$ / $C_{\text{max}}$    & A strategy's cooperating rate divided by the maximum \\
$C_r$ / $C_{\text{min}}$    & A strategy's cooperating rate divided by the minimum \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median \\
$C_r$ / $C_{\text{mean}}$   & A strategy's cooperating rate divided by the mean \\
$C_r$                       & The cooperating ratio of a strategy \\
$CC$ to $C$ rate            & The probability a strategy will cooperate after a mutual cooperation \\
$CD$ to $C$ rate            & The probability a strategy will cooperate after being betrayed by the opponent \\
$DC$ to $C$ rate            & The probability a strategy will cooperate after betraying the opponent\\
$DD$ to $C$ rate            & The probability a strategy will cooperate after a mutual defection\\
$p_n$                       & The probability of a player's action being flip at each interaction \\
$n$                         & The number of turns \\
$p_e$                       & The probability of a match ending in the next turn \\
$N$                         & The number of strategies in the tournament \\
$k$                         & The number that a given tournament is repeated \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The features which are included in the performance evaluation analysis.}\label{table:parameters_summary}
\end{table}

>>>>

For the purposes of this manuscript a data set containing results of IPD
tournaments has been generated and is available at~\cite{data}. This was done using the
open source package Axelrod-Python library~\cite{axelrodproject} (APL), and more specifically,
version 3.0.0. APL allows for different types of IPD computer
tournaments to be simulated whilst containing a list of over 180 strategies.
Most of these are strategies described in the literature with a few exceptions
being strategies that have been contributed specifically to the package. This
paper makes use of \numberofstrategies strategies implemented in version 3.0.0. A
list of the strategies is given in the Appendix~\ref{app:list_of_players}.
Although APL features several tournament types, this work considers
standard, noisy, probabilistic ending and noisy probabilistic ending
tournaments.

@@

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014}, which are
strategies based on finite state automata introduced by Daniel Ashlock and Wendy
Ashlock. These strategies have been evolved using reinforcement learning, however,
there were trained to maximise their payoffs in tournaments with fixed turns
(150 specifically) and not in probabilistic ending ones. In probabilistic ending
tournaments it appears that the top ranks are mostly occupied by defecting
strategies. These include Better and Better, Gradual Killer, Hard Prober (all
from ~\cite{prison}), Bully (Reverse Tit For Tat)~\cite{Nachbar1992} and
Defector. Thus, it's surprising that EasyGo and Fool Me Forever which are
strategies that will defect until their opponent defects, then they will cooperate
until the end, ranked $14^{\text{th}}$ and $15^{\text{th}}$. Upon inspection,
it was found that they are actually the same strategy. This was not known to
the authors at the time of data collection. Figure~\ref{fig:probend_results}
verifies that their performance is the same. Both strategies have repeatedly
ranked highly and there are cases for which they were the winners of the
tournament.

@@

The top ranks include strategies that have been previously mentioned. The
set of Retaliate strategies occupy the top spots followed by BackStabber
and DoubleCrosser. The distributions of the Retaliate strategies have no
statistical difference. Thus, in an IPD tournament where the type is not
specified, playing as any of the Retaliate strategies will have the result.
DoubleCrosser performed well in standard tournaments and the
strategy is just an extension of BackStabber. It should be noted that these
strategies can be characterised as "cheaters". The source code of the strategies
allows them to known the number of turns in a match (if they are specified).
PSO Gambler and Evolved HMM 5 are
trained strategies introduced in~\cite{Harper2017} and Nice Meta Winner and NMWE
Memory One are strategies based on teams. Grudger is a strategy from Axelrod's
original tournament and Forgetful Fool Me Once is based on the same approach as
Grudger. Overall the top 15 strategies are fundamentally different. Some are cheaters,
some are complex, others are simple deterministic strategies and strategies based
on teams. The results of \numberofalltournaments tournaments used in this work imply the following:
they is not a single type of strategy performs well in any IPD interaction.

@@

Compared to standard tournaments, in both noisy and in probabilistic ending
tournaments the higher the rates of cooperation the lower a strategy's success
and median score. A strategy would want to cooperate less than both
the mean and median cooperator in such settings. In probabilistic ending
tournaments the correlations coefficients have a larger values, indicating a
stronger effect. Thus a strategy will be punished more by its cooperative
behaviour in probabilistic ending environments, this was seen in Section~\ref{section:evaluation_of_performance}
as well. The distributions of the $C_r$ of the winners in
both tournaments is given by Figure~\ref{fig:c_r_distributions}. It confirms
that the winners in noisy tournaments cooperated less than 35\% of the times
and in probabilistic ending tournaments less than 10\%.
In noisy probabilistic ending tournaments and in over all the tournaments' results,
the only features that had a moderate affect are $C_r/C_{\text{mean}},
C_r/C_{\text{max}}$ and $C_r$. In such environments cooperative behaviour
appears to be punished by not as much as in noisy and probabilistic ending
tournaments.

@@

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.9\textwidth}{!}{
        \begin{tabular}{lccccc}
    \toprule
    Tournament type & Clustering Approach & Number of clusters & $R^2$ training data &  $R^2$ test data  & $R^2$ OOB score\\
    \midrule
    standard  & Approach 1  & 2 & 0.998831  & 0.987041    & 0.983708 \\
              & Approach 2  & 2 & 0.998643  & 0.978626    & 0.969202 \\
              & Approach 3  & 2 & 0.998417  & 0.985217    & 0.976538 \\
              & Approach 4  & 2 & 0.998794  & 0.990677    & 0.982959 \\
    \midrule
    noisy     & Approach 1  & 2 & 0.997786  & 0.972229    & 0.968332\\
              & Approach 2  & 2 & 0.997442  & 0.963254    & 0.955219\\
              & Approach 3  & 2 & 0.997152  & 0.953164    & 0.940528\\
              & Approach 4  & 3 & 0.996923  & 0.950728    & 0.935444\\
    \midrule
    probabilistic ending & Approach 1 & 2 & 0.997909   & 0.981490  & 0.978120 \\
                         & Approach 2 & 2 & 0.997883   & 0.973492  & 0.967150 \\
                         & Approach 3 & 2 & 0.990448   & 0.890068  & 0.875822 \\
                         & Approach 4 & 2 & 0.999636   & 0.995183  & 0.992809 \\
    \midrule
    noisy probabilistic ending & Approach 1  & 2 & 0.995347 & 0.957846   & 0.952353\\
                               & Approach 2  & 2 & 0.992813 & 0.909346   & 0.898613\\
                               & Approach 3  & 2 & 0.990579 & 0.824794   & 0.806540\\
                               & Approach 4  & 4 & 0.989465 & 0.841652   & 0.824052\\
    \midrule
    over \numberofalltournaments tournaments   & Approach 1 & 2 & 0.997271& 0.972914& 0.969198 \\
                                               & Approach 2 & 2 & 0.996323& 0.951194& 0.940563 \\
                                               & Approach 3 & 2 & 0.993707& 0.906941& 0.891532 \\
                                               & Approach 4 & 3 & 0.993556& 0.913335& 0.898453 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Accuracy metrics for random forest models.}
    \label{table:accuracy_random_forest}
\end{table}

@@

The \numberofstrategies strategies used in this manuscript have been mainly for
the literature, and they have been accessible due to an open source software
called the Axelrod-Python library. The software was used to generate a total of
\numberofalltournaments computer tournaments results with different number of
strategies and different participants each time. The data collection was
described in Section~\ref{section:data_collection}. In
Section~\ref{section:top_performances}, the tournaments results were used to
present the top performances. The data set contained results from four different
settings, and these were also studied individually. In standard tournaments
complex strategies trained using reinforcement learning ranked in the top spots.
Some of these strategies ranked again in the top spots in probabilistic
ending tournaments when a \(p_e\) of less 0.1 was considered. In probabilistic
ending tournaments \(p_e\) was designed to vary between 0 and 1. It was demonstrated
that for values larger than 0.1, as stated in the Folk Theorem, defecting strategies
were winning the tournaments because there was a high likelihood of the game
ending in the next turn. In tournaments with noise the median ranks of the top
15 strategies had the highest values and the \(r\) distributions were bimodal.
The top rank strategies were performing both well and bad, and this indicates
that in noisy tournaments there are not strategies that can guarantee winning.
Overall, the top ranked strategies differed from one tournament type to
another and the mechanism behind the winning strategies were all different.
Even strategies designed to perform well in one setting were demonstrated to be
better in others.

Section~\ref{section:evaluation_of_performance} covered an analysis of
performance based on several features associated with a strategy and with the
environments in which it was competing. The results of this analysis showed that a
strategy's characteristics such as whether or not it's stochastic, and the information it
used regarding the game had no effect on the strategy's success. The most
important features have been those that compared the strategy's behaviour to its
environment. The cooperating ratio of the strategy compared to the mean and
median cooperator was highlighted as the most important feature in the analysis.
More specifically, if a strategy were to enter a tournament with a theory of
mind of its environment it would choose to be the median cooperator in standard
tournaments, to cooperate 10\% of the time the median cooperator did in probabilistic ending tournaments and 
60\% in noisy and noisy probabilistic
tournaments. Lastly, if a strategy was aware of the opponents but not of the
setting of the tournament, a strategy would be more likely to be successful if
it were to identify the median cooperator and cooperated 67\% of the times that
they did.

@@

\begin{table}[h]
    \begin{center}
    \resizebox{.7\textwidth}{!}{
    \begin{tabular}{llc}
    \toprule
feature & feature explanation \\
 \midrule
stochastic                  & If a strategy is stochastic \\
makes use of game           & If a strategy makes used of the game information  \\
makes use of length         & If a strategy makes used of the number of turns \\
memory usage                & The memory size of a strategy divided by the number of turns \\
SSE                         & A measure of how far a strategy is from extortionate behaviour \\
$C_{\text{max}}$          & The biggest cooperating rate in the tournament \\
$C_{\text{min}}$          & The smallest cooperating rate in the tournament \\
$C_{\text{median}}$       & The median cooperating rate in the tournament \\
$C_{\text{mean}}$         & The mean cooperating rate in the tournament \\
$C_r$ / $C_{\text{max}}$    & A strategy's cooperating rate divided by the maximum \\
$C_r$ / $C_{\text{min}}$    & A strategy's cooperating rate divided by the minimum \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median \\
$C_r$ / $C_{\text{mean}}$   & A strategy's cooperating rate divided by the mean \\
$C_r$                       & The cooperating ratio of a strategy \\
$CC$ to $C$ rate            & The probability a strategy will cooperate after a mutual cooperation \\
$CD$ to $C$ rate            & The probability a strategy will cooperate after being betrayed by the opponent \\
$DC$ to $C$ rate            & The probability a strategy will cooperate after betraying the opponent\\
$DD$ to $C$ rate            & The probability a strategy will cooperate after a mutual defection\\
$p_n$                       & The probability of a player's action being flipped at each interaction \\
$n$                         & The number of turns \\
$p_e$                       & The probability of a match ending in the next turn \\
$N$                         & The number of strategies in the tournament \\
$k$                         & The number that a given tournament is repeated \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The features which are included in the performance evaluation analysis.}\label{table:parameters_summary}
\end{table}
